---
title: "Bakery Market Basket Analysis"
author: "Stone Shi"
date: "11/18/2019"
output:
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Collect the data
```{r, message = FALSE}
library(arules)
library(tidyverse)

bread <- read.transactions("BreadBasket_DMS.csv", cols = c(3,4), sep = ",", format = "single", quote = "\"", rm.duplicates = TRUE)

```

# 2. Exploratory Data Analysis
## 2.1 Basic Summary Output
Let's take a look at the summary stats for the dataset. Through the summary output, we can see that there are in total **6614** transactions and **104** items in my dataset. The top 5 frequent items are: *Coffee*, *Bread*, *Tea*, *Cake* and *Pastry*. The length distribution tells us there are 2556 transactions contain one item, 2154 transactions contain two items, 1078 transactions contain 3 items, etc. The maximum count of item in a single transaction is 10.

```{r}
summary(bread)
```

## 2.2 Transactions Exploration
```{r}
image(sample(bread,100))
```

Let's look at our first five transactions. 
```{r}
inspect(bread[1:100])
```

Let's get to know the frequency of our first five items from our transactions.

```{r}
itemFrequency(bread[,1:5])
```

Then we can also utilize the *itemFrequencyPlot()* function to visualize the frequency of our items.

Let's produce a histogram of items with at least 5 percent support.
```{r}
itemFrequencyPlot(bread, support = 0.05)
```

We can also produce a histogram for the top 20 frequent item sets.
```{r}
itemFrequencyPlot(bread, topN = 20)
```

#3. Market Basket Analysis
The *arules* pakcage in R has a limited number of functions for exploring the data. So in order to continue our market basket analysis, we need to create a dataframe of the items and their purchase frequency. Let's take a look at the whole itemFrequency.

```{r}
itemFrequency(bread)
```

Now, we can convert the data to a data frame.

```{r}
bakery.frequency <-
  data.frame(
    Items = names(itemFrequency(bread)),
    Frequency = itemFrequency(bread),
    row.names = NULL
  )

head(bakery.frequency)
```


> 10 most frequently bought items at the store.

```{r}
bakery.frequency %>% 
  arrange(desc(Frequency)) %>% 
  slice(1:20) %>% 
  ggplot() +
  geom_col(aes(x = reorder(Items, Frequency), y = Frequency), fill = "lightblue", color = "black")+
  labs(x = "Items") +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none")
        
```

> 10 least frequently bought items at the store.

```{r}
bakery.frequency %>% 
  arrange(Frequency) %>% 
  slice(1:20) %>% 
  ggplot() +
  geom_col(aes(x = reorder(Items, -Frequency), y = Frequency), fill = "#f6abb6", color = "black")+
  labs(x = "Items") +
  coord_flip() +
  theme_minimal() +
  theme(legend.position = "none")
```

# 3. Train the model

I will use **apriori()** function to generate some rules. 

```{r}
bakery_rules <- apriori(bread, parameter = list(support = 0.01, confidence = 0.5, minlen = 2))
```


# 4. "Evaluate" the Model's Performance.
## 4.1 Understand the rules
Let's first get the summary of the bakery association rules.
```{r}
summary(bakery_rules)
```


Then we can take a look at the rules we have:
```{r}
inspect(bakery_rules)
```

From the rules we have above, we can convey some key information:

* 59% of the customers who bought a spanish brunch also bought a coffee.
* 70% of the customers who bought a toast also bought a coffee.

Looks like most of the customers who come to the bakery will buy a cup of coffee!

## 4.2. Visualize the rules

We can also **visualize** our rules.

```{r}
library(arulesViz)
plot(bakery_rules, measure = c("support", "lift"), shading = "confidence")
```

We can use another visualization to represent our rules.

```{r}
plot(bakery_rules, method = "graph")
```

# 5. "Improve" the Model's Performance

I saw the item "NONE" a lot of times through our analysis, so I want to eliminate it.
```{r}
bakery_rules %>% 
  subset(!lhs %in% "NONE") %>% 
  plot(., method = "graph")
```

Now our rules look good! 